---
title: "`r paste('&nbsp; ', params$opt$project)`"
vignette: >
  %\VignetteIndexEntry{CellExpress report}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  html_document:
    css: style.css
    fig_caption: true
    df_print: kable
    toc: true
    toc_float: true
    number_sections: true
params:
    snapshot_file: "snapshot.json"
    opt: NULL
    disease_id: NULL
    disease_label: NULL
    tissue_id: NULL
    tissue_label: NULL
---

<style> 
  #header { 
    background: -moz-linear-gradient(left, rgba(120,21,79,1) 0%, rgba(120,21,79,1) 75%); /* FF3.6-15 */
    ackground: -webkit-linear-gradient(left, rgba(120,21,79,1) 0%,rgba(120,21,79,1) 75%); /* Chrome10-25,Safari5.1-6 */
    background: linear-gradient(to right, rgba(120,21,79,1) 0%,rgba(120,21,79,1) 75%); /* W3C, IE10+, FF16+, Chrome26+, Opera12+, Safari7+ */
    filter: progid:DXImageTransform.Microsoft.gradient( startColorstr='#6f13dd', endColorstr='#007db9e8',GradientType=1 ); /* IE6-9 */
    color: white;
    height: 110px;
  }
</style>

<!-- create a tooltip that appears when a user hovers over a "Read me" text --> 
<style>
/* Tooltip container */
.tooltip-custom {
  position: relative;
  display: inline-block;
  cursor: help;
}

.tooltip-custom-label {
  color: #1f77b4;
}

/* Tooltip text */
.tooltip-custom::after {
  content: attr(data-tooltip); 
  visibility: hidden;
  background-color: #FFB3E2;
  color: #000000;
  text-align: left;
  padding: 5px 10px;
  border-radius: 5px;
  position: absolute;
  width: max-content;
  max-width: 500px;
  bottom: 120%;
  left: 50%;
  transform: translateX(-50%);
  opacity: 0;
  transition: opacity 0.3s ease-in-out;
  font-size: 12px;
}

/* Show the tooltip when hovering */
.tooltip-custom:hover::after {
  visibility: visible;
  opacity: 1;
}
</style>

<!-- *** --> 

```{r html, echo=FALSE}
htmltools::img(src = knitr::image_uri("cellexpress_logo_report.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:22px; right:40px; width: 24%; height: 106px;'
               )
```

```{r setup, include=FALSE}
# Load required libraries
packages <- c("jsonlite", "ggplot2", "dplyr", "gtools", "gridtext", "grid", "kableExtra", "DT", 
              "scales","RColorBrewer", "plotly", "visNetwork", "ggrepel", "gtools", "viridis")
lapply(packages, library, character.only = TRUE)

# source helper functions
source(file.path(params$opt$cellexpress_path, "report_helper.R"))

# Read the snapshot data from the JSON file provided by Python
rprt <- fromJSON(params$snapshot_file)

# Extract summary data
metadata_df <- dplyr::bind_rows(rprt$metadata_df)
summary_df <- dplyr::bind_rows(rprt$summary_df)
umi_per_cell <- dplyr::bind_rows(rprt$umi_per_cell)
gene_per_cell <- dplyr::bind_rows(rprt$gene_per_cell)
mt_per_cell <- dplyr::bind_rows(rprt$mt_per_cell)

# Extract QC metric
db_plots <- dplyr::bind_rows(rprt$db_plots)
filters <- dplyr::bind_rows(rprt$filters)
bc.qc.raw <- dplyr::bind_rows(rprt$bc_qc_raw) %>%
    tibble::column_to_rownames("index")
bc.qc.ji <- dplyr::bind_rows(rprt$bc_qc_ji) %>%
    tibble::column_to_rownames("index")

# get colors
set.seed(123)  
smpl_colrs <- color_helper(set.names=sort(unique(metadata_df$sample_id)), panel="Set1") 
smpl_colrs <- setNames(as.vector(smpl_colrs)[sample(1:length(smpl_colrs), size = length(smpl_colrs), replace = FALSE,)], names(smpl_colrs))
```

# ExpressSummary 

<u class="tooltip-custom tooltip-custom-label" data-tooltip="The **ExpressSummary** section presents key experimental context and QC numerical values: 
Total number of cells (input data), displayed in grey, representing the initial number of cells before any filtering.
Number of cells after QC, highlighted in green, indicating the number of cells retained post-quality control.
Mean number of UMIs per cell, shown in orange, representing the average transcript counts per cell.
Mean number of genes per cell, marked in blue, summarizing the average gene detection per cell.">Read me</u>

## Sample & Experimental Context

```{r test, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.height=1.0, fig.width=10.0, results='asis'}
data.frame(`Species` = ifelse(params$opt$species == "hs", "human", "mouse"), 
           `Disease` = params$disease_label, 
           `Tissue`= params$tissue_label, 
           check.names = FALSE) %>%
  kableExtra::kable("html", align = "c", escape = FALSE, # escape = FALSE allows HTML rendering
                    col.names = c("Species", "Disease", "Disease ID", "Tissue", "Tissue ID")) %>% 
  kableExtra::kable_classic(full_width = TRUE, html_font = "Cambria") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                            full_width = TRUE, fixed_thead = TRUE, position = "center") %>%
  kableExtra::row_spec(0, background = "#e7298a", color = "white", bold = TRUE, font_size = 18) %>%  # Header color
  kableExtra::row_spec(1, background = "#fce5ec", color = "#333333", font_size = 17)  # Row color
```

- - -

# ExpressQC

<u class="tooltip-custom tooltip-custom-label" data-tooltip="The **ExpressQC** section provides a comprehensive 
overview of sample-level quality metrics. It includes tabulated summaries of metadata, cell and gene counts 
before QC, as well as quantiles of UMIs, genes, and mitochondrial content per cell. Interactive density plots show 
the distribution of key QC metrics across cells. Barcode overlap metrics assess potential sample cross-contamination.">Read me</u>

```{r liberal-filter-comment, echo=FALSE, results='asis'}
cat("<div style='padding: 10px; background-color: #f9f9f9; border-left: 5px solid #007bff;'>")
cat("<strong>Note:</strong> This analysis was run in <em>QC-only mode</em>. As such, minimal filtering thresholds were applied to retain as many cells and genes as possible while only removing entries with zero UMIs.")
cat("</div>")
```

## Sample-level Characteristics and Quality Control Metrics {.tabset .tabset-fade .tabset-dropdown}

### Metadata

```{r metadata, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.align = 'center'}
metadata_df %>%
  dplyr::mutate(sample_id = factor(sample_id, levels = sample_id[mixedorder(sample_id)], ordered = TRUE)) %>%
  dplyr::arrange(sample_id) %>%
  kable() %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "200px")
```

### Cells & Genes

```{r cells_genes, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.align = 'center'}
summary_df %>%
  dplyr::rename(pre_qc_cell = cells,
                pre_qc_gene = genes) %>%
  dplyr::mutate(pre_qc_cell = scales::comma(pre_qc_cell, accuracy = 1),
         pre_qc_gene = scales::comma(pre_qc_gene, accuracy = 1)) %>%
  dplyr::select(-type) %>%
  dplyr::mutate(sample_id = factor(sample_id, levels = sample_id[mixedorder(sample_id)], ordered = TRUE)) %>%
  dplyr::arrange(sample_id) %>%
  kable() %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "200px")
```

### UMI per cell

```{r umi, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.align = 'center'}
umi_per_cell %>%
  dplyr::mutate(across(where(is.numeric), ~ scales::comma(., accuracy = 1))) %>%
  kable() %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "200px")

```

### Gene per cell

```{r gene, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.align = 'center'}
gene_per_cell %>%
  dplyr::mutate(across(where(is.numeric), ~ scales::comma(., accuracy = 1))) %>%
  kable() %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "200px")
```

### Mitochondrial (%)

```{r mt, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.align = 'center'}
mt_per_cell %>%
  dplyr::mutate(across(where(is.numeric), ~ sprintf("%.1f%%", .))) %>%
  kable() %>%
  kable_classic(full_width = F, html_font = "Cambria") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                full_width = F, fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "200px")
```

## {-}

- - -

```{r pt_density, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
  # ===================================
  Q <- list()
  for (tp in unique(db_plots$type)) {
    p <- ggplot() +
      getBaseTheme() +
      theme(strip.text = element_text(size = 20, face = "bold"),
            axis.title = element_text(size = 12, face = "bold"),
            axis.text = element_text(size = 10, face = "bold"),
            legend.title = element_text(size=13),
            legend.text = element_text(size=11)) +
      scale_x_continuous(trans = scales::log10_trans(),
                         breaks = scales::trans_breaks("log10", function(x) 10^x),
                         labels = scales::trans_format("log10", scales::math_format(10^.x))) +
      xlab("10^x") + ylab("Density") +
      geom_density(data = filter(db_plots, type == tp),
                   aes(x = feature, color = sample_id, feature = feature), 
                   show.legend = FALSE, na.rm = TRUE) +
      scale_color_manual(values = smpl_colrs) +
      guides(fill = "none", color = "none")
    Q[[length(Q) + 1]] <- p  
  }
  # ===================================
  # Convert long-format db_plots into a wide-format dataframe (fobj)
  df <- db_plots %>%
    tidyr::pivot_wider(id_cols = c(barcode, sample_id), names_from = type, values_from = feature)
  pqc <- ggplot(df,
              aes(x = nFeature_RNA, y = nCount_RNA, color = sample_id)) +
        getBaseTheme() +
        theme(strip.text = element_text(size = 20, face = "bold"),
              axis.title = element_text(size = 12, face = "bold"),
              axis.text = element_text(size = 10, face = "bold"),
              legend.title = element_blank(),
              legend.position = "right",
              legend.text = element_text(size=11)) +
        # geom_point(alpha = 0.4, size = 1) +
        geom_smooth(aes(group = sample_id), 
                    method = "loess", se = FALSE, size = 0.5, fullrange = FALSE) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
        scale_x_continuous(trans = scales::log10_trans(),
                            breaks = scales::trans_breaks("log10", function(x) 10^x),
                            labels = scales::trans_format("log10", scales::math_format(10^.x))) +
        scale_y_continuous(trans = scales::log10_trans(),
                            breaks = scales::trans_breaks("log10", function(x) 10^x),
                            labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
        xlab("Genes detected [10^x]") + ylab("UMI counts [10^x]") +
        scale_color_manual(values = smpl_colrs) +
        guides(fill = "none", color = "none")
  # ===================================
```

## Distribution Plots for Quality Control Metrics {.tabset .tabset-fade .tabset-dropdown}

### UMI per cell distribution

<center>
```{r q1, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.dim=c(5.5, 4.5), fig.align = 'center'}
ggplotly(Q[[1]], tooltip = c("sample_id", "feature"))
```
<center>

### Gene per cell distribution

<center>
```{r q2, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.dim=c(5.5, 4.5), fig.align = 'center'}
ggplotly(Q[[2]], tooltip = c("sample_id", "feature"))
```
<center>

### Mitochondrial (%) per cell distribution

<center>
```{r q3, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.dim=c(5.5, 4.5), fig.align = 'center'}
ggplotly(Q[[3]], tooltip = c("sample_id", "feature"))
```
<center>

### Total UMI Counts vs. Genes Detected per cell

The gray dashed line indicates the identity line (y = x)

<center>
```{r pqc, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.dim=c(5.75, 4.5), fig.align = 'center'}
# plot(pqc)
ggplotly(pqc, tooltip = c("sample_id", "nFeature_RNA", "nCount_RNA"))
```
<center>

## {-}

- - -

## Barcodes Contamination {.tabset}

### Raw count

Number of barcodes shared between pairs of samples post-QC.

```{r bcn_raw, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.dim=c(5.5, 5), fig.align = 'center'}
if (length(bc.qc.raw) != 0) {
  bc.qc.raw %>%
    kable() %>%
    kable_classic(full_width = F, html_font = "Cambria") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, fixed_thead = TRUE) %>%
    scroll_box(width = "100%", height = "200px")  
} else {
  print("No overlaps among samples' barcodes.")
}
```

### Jaccard Index

Fraction (%) of barcodes shared between pairs of samples pre-QC.

```{r bcn_pct, eval=TRUE, warning=FALSE, message=FALSE, echo=FALSE, fig.dim=c(5.5, 5), fig.align = 'center'}
if (length(bc.qc.ji) != 0) {
  bc.qc.ji %>%
    kable() %>%
    kable_classic(full_width = F, html_font = "Cambria") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                  full_width = F, fixed_thead = TRUE) %>%
    scroll_box(width = "100%", height = "200px")  
} else {
  print("No overlaps among samples' barcodes.")
}
```

## {-}

- - -